Combinatorics of the geometry of Wilson loop diagrams II: Grassmann necklaces, dimensions, and denominators

### 2019-04-26 GWLDCII.tex
copyedits to section 1 and 2 of paper 2

This paper is the second in a two part series investigating the combinatorics and geometry underlying SYM $N=4$ theory. This series lays out several results about the relationship between the Wilson loop diagrams and the positroid cells that give a CW complex structure to the positive Grassmanian. This paper is concerned with identifying which positroid cells correspond to which Wilson loop diagrams, and how this translates to a geometric tiling of a subspace of the positive Grassmannians. Wilson loop diagrams associate a volume form to a positroid cell. In this paper, we give an algorithm to identify which cell corresponds to which diagram, via Grassmann Necklaces. Using the Le diagram representation of positroid cells, we prove results about the dimensions of these cells. Finally, we state some results about the structures of the volume forms defined on these positroid cells by the Wilson loop diagrams.

In contrast, the first paper in this series is about combinatorial and matroidal properties of the Wilson loop diagrams. The main focus of that paper was to enumerate the positroid cells that appeared in the image of the association between diagrams and cells, as well as the size of the number of Wilson loop diagrams in the pre-image of any positroid cell under this association.s is given by counting the directions of associahedrons.

In recent years, there has been a some interest in understanding the geometry and combinatorics underlying the field theory SYM $N=4$ \cite{wilsonloop, Arkani-Hamed:2013jha, Amplituhedronsquared, AmplituhedronDecomposition, galashinlam}. This started with the observation BCFW diagrams, which represent the on shell interactions of this theory correspond to plabic graphs \cite{GrassmannAmplitudebook}. Plabic graphs are one of the many equivalent ways to identify positroid cells of the positie Grassmannians \cite{Postnikov}. This work led to the representation of the on shell integrals associated this theory in terms of a geometric space embedded in $\Gr(k,n)$, called the Amplituhedron \cite{Arkani-Hamed:2013jha, Arkani-Hamed:2013kca}. More over, inegrals associated to the BCFW diagrams define a volume form on the Amplituhedron. Since then, there has been significant work to understand the volume form and its physical implications \cite{galashinlam, UnwindingAmplituhedron}.

Meanwhile, a different body of work has studied SYM $N=4$ theory from the point of view of Feynman integrals in twistor space \cite{Adamo:2012xe, Adamo:2011pv,  Boels:2007qn, Bullimore:2010pj}. These integrals are calculated via holomorphic Wilson loops. As with the Amplituhedron, each Wilson loop represents the sum of a family of $N^kMHV$ Feynman diagrams. The sum of the corresponding sum of Feynman integrals associate a volume form to space \cite{Hodges:2012ym}. The first author and Amat uncovered a connection between the Feynman integrals developed in this literature and positroids, defined  as matroids that can be realized as an element of $\Gr(k,n)$ \cite{wilsonloop}, while other work has shown a connection between the space parametrized by Wilson loop diagrams and the Amplituhedron \cite{casestudy, Amplituhedronsquared}.

In the above work and elsewhere, it is clear that while the space parametrized by the Wilson loop diagrams is related to the Amplituhedron. They are not the same. For in stance, \cite{non-orientable} shows that the space parametrized by the Wilson loop diagrams can be seen as a non-orientable vector bundle over some submanifold of $\Gr(k,n)$, while it is conjectured that the Amplituhedron is orientable. Furthermore, the cells associated to the Wilson loop diagrams are different from the cells associated to BCFW diagrams, as one sees from counting arguments in comparing the results of \cite{AmplituhedronDecomposition} and \cite{casestudy}. This is not too alarming, as one expects space parametrized by the Wilson loop diagrams to project down to something close to the Amplituhedron \cite{wilsonloop, Amplituhedronsquared}.

This paper continues this work by identifying which positroid cells are associated to which Wilson loop diagram. Where the BCFW diagrams identify positroid cells via plabic graphs, in this paper, we give an algorithm to read off of a Wilson loop diagram the associated Grassmann Necklace. Furthermore, while the BCFW cells are all 4k dimensional, we show that the positroid cells associated to Wilson loop diagrams are $3k$ dimensional. Again, this is not alarming, as \cite{non-orientable} has shown that the actual geometric space parametrized by the Wilson loop diagrams is a non-orientable $k$ dimensional bundle over this $3k$ dimensional manifold in $\Gr(k,n)$.

Finally, when computing a finite volume to a geometric object defined by the Wilson loop diagrams, one must ensure at any singularities cancel. This has been expicitly shown in a few cases \cite{casestudy, HeslopStewart}, but it has been hard to prove exactly due to the non-one to one or onto relationship between Wilson loop diagrams and $3k$ dimensional positroid cells of $\Gr(k,n)$. In this paper, we give a characterisation of the denominator of the Feynman integral defined by a Wilson loop diagram in terms of the algorithm that associated the same diagram to a Grassman Necklace.

@@

Section~\ref{section general background} gives the required background for Wilson loop diagrams (subsection~\ref{section WLD background}) and for positroids and some of their many combinatorial versions (subsection~\ref{sec:positroid background}).

Section~\ref{sec GN algorithm} proceeds to give an algorithm to go directly from a Wilson loop diagram to its Grassmann necklace (Algorithm~\ref{alg:put GN on WLD} and the proof that it is correct in Theorem~\ref{res:alg gives GN}).  In view of the known combinatorial equivalences, it was already possible to get the Grassmann necklace of a Wilson loop diagram, but it involved going through the matrix $C(W)$ and the positroid as a matroid.  The direct algorithm shows us directly how each propagator contributes to each element of the Grassmann necklace.  This idea is crucial for all the arguments of the present paper.  Along with the algorithm, Section~\ref{sec GN algorithm} contains two key techincal lemmas, Lemma~\ref{lem sian}, the focus of subsection~\ref{sec:propagator configs}, and the cyclic interval lemma, Lemma~\ref{vertex cyclic int lem} in subsection~\ref{sec:GN alg}, which leads into the proof of Theorem~\ref{res:alg gives GN} as well as being important in subsequent sections.

Section~\ref{sec dim} proceeds to show that the dimension of the positroid cell defined by a Wilson loop diagram has dimension three times the number of propagators, Theorem~\ref{thm dim}.  The proof here is very explicit, keeping track of how a propagator contributes $+$s to the Le diagram, and correspondingly somewhat technical.  Section~\ref{sec poles} characterizes the denominator of the integral of a Wilson loop diagram in terms of how propagators contribute to the Grassmann necklace.  This result is also given by an algorithm, Algorithm~\ref{alg WLD to denom via GN}, and theorem, Theorem~\ref{thm denom}.

@@

Generally speaking the propagators are undirected, so $(i,j)$ and $(j,i)$ represent the same propagator. As a general convention, we write $p = (i,j)$ with  $i < j$ relative to the first vertex. However, in Section~\ref{sec:propagator configs} we will need to give the propagators a direction (in order to refer to the regions ``inside'' and ``outside'' the propagator); in this setting, we write $(p,<_i)$ to indicate that $p$ is directed from $i$ to $j$, and $(p,<_j)$ for the same propagator directed from $j$ to $i$.

@@

We first introduce some definitions that will allow us to talk about subsets of the vertices of $W$, subsets of the propagators of $W$, and what it means to be a subdiagram of $W$.

@@

For our purposes, a {\bf positroid} is a matroid $M$ (with ground set $[n]$ and bases $\cB \subseteq \binom{[n]}{k}$) which can be represented by an element of the totally nonnegative Grassmannian $\Gr(k,n)$. In other words, there exists a full-rank $k\times n$ real matrix whose $k\times k$ minors are all nonnegative and such that the minor $[[k]|J]$ is positive if and only if $J \in \cB$.

The interested reader is referred to \cite{Postnikov} for an introduction to the geometry and combinatorics of positroid cells in the totally nonnegative Grassmannian. Postnikov shows in \cite{Postnikov} that the positroids of $\Gr(k,n)$ are indexed by many different collections of objects, each with their own advantages and disadvantages. The two most suited to our purposes are Grassmann necklaces and Le diagrams, which we introduce below. In order to do so, we first need some preliminary definitions.

@@

By \cite[Theorem 17.1]{Postnikov}, the Grassmann necklaces of type $(k,n)$ are in 1-1 correspondence with the positroid cells in $\Gr(k,n)$.  This correspondence is given explicitly in \cite[Theorem 8]{Oh}: if $(I_1, \dots, I_n)$ is the Grassmann necklace associated to a positroid $M = ([n],\cB)$, then the bases of $M$ are exactly
\[\cB = \left\{J \in \binom{[n]}{k}\ |\ I_i \gale{i} J \ \forall i \in [n]\right\}.\]
Thus the Grassmann necklace is well suited to testing whether a given $k$-set is a basis for $M$, and for generating a list of all bases of $M$.

@@

An algorithm for constructing the Le diagram associated to a Grassmann necklace was given by Agarwala and Fryer in \cite{reversingOh}. Since we will make use of this algorithm in Section~\ref{sec dim} below, we summarise the process here.
\begin{algorithm}\label{alg:GN to Le} \ \cite[Algorithm 2]{reversingOh}
Let $(I_1,\dots,I_n)$ be a Grassmann necklace of type $(k,n)$. Within a $k \times(n-k)$ square, draw the Young diagram whose rows are labelled by $I_1$ (as per the convention above).

@@

\begin{thm} \label{thm WLD defines matroid} \cite[Theorem 3.6]{wilsonloop} The independent sets of the matroid $M(W)$ associated to an admissible Wilson loop diagram $W = (\cP,[n])$ and realized by $C(W)$ are exactly those subsets $V \subseteq [n]$ such that $\nexists U\subseteq V$ satisfying $|\Prop(U)| < |U|$. \end{thm}
In other words, the independent sets of $M(W)$ correspond to the sets of vertices in $W$ such that no subset supports fewer propagators than the vertices it contains.

By studying $M(W)$ in terms of the behavior of the propagators in $W$, Agarwala and Amat show that $M(W)$ is in fact a positroid whenever $W$ is admissible \cite[Corollary 3.39]{wilsonloop}.

@@

\begin{dfn}\label{props inside p}
Let $W = (\cP, [n])$ be a weakly admissible Wilson loop diagram, with $p \in \cP$ supported on edges $i$ and $j$. Write $(p, <_i)$ to be the same propagator with a direction induced by the linear ordering $<_i$. Then the set of propagators {\bf inside} $(p,<_i)$ is defined to be
\[\cP_{in}(p,<_i) = \{ q \in \cP \ |\ (q,<_i) = (k,l) \text{ and }i \leq_i k <_i l \leq_i j \} \]
and the set of propagators {\bf outside} $(p,<_i)$ is
\[\cP_{out}(p,<_i) = \cP\setminus \cP_{in}(p,<_i).\]
\end{dfn}

>>>>

This paper is the second in a two part series investigating the combinatorics and geometry underlying SYM $N=4$ theory. This series lays out several results about the relationship between the Wilson loop diagrams and the positroid cells that give a CW complex structure to the positive Grassmanian. This paper is concerned with identifying which positroid cells correspond to which Wilson loop diagrams, and how this translates to a geometric tiling of a subspace of the positive Grassmannians.

% Wilson loop diagrams associate a volume form to a positroid cell.

% In this paper, we give an algorithm to identify which cell corresponds to which diagram, via Grassmann Necklaces. Using the Le diagram representation of positroid cells, we prove results about the dimensions of these cells. Finally, we state some results about the structures of the volume forms defined on these positroid cells by the Wilson loop diagrams.

In contrast, the first paper in this series focuses on combinatorial and matroidal properties of the Wilson loop diagrams. The main focus of that paper was to enumerate the positroid cells that could be associated to Wilson loop diagrams with a fixed number of vertices and propagators, as well as counting the number of Wilson loop diagrams in the pre-image of any positroid cell under this association.

In recent years, there has been significant interest in understanding the geometry and combinatorics underlying the field theory SYM $N=4$ \cite{wilsonloop, Arkani-Hamed:2013jha, Amplituhedronsquared, galashinlam, AmplituhedronDecomposition}. This started with the observation that BCFW diagrams, which represent the on shell interactions of this theory, correspond to plabic graphs \cite{GrassmannAmplitudebook}. Plabic graphs were introduced by Postnikov in \cite{Postnikov}, and provide one of the many equivalent ways to classify the positroid cells of the positive real Grassmannian $\Gr(k,n)$. This led to the representation of the on shell integrals associated to this theory in terms of a geometric space (the Amplituhedron) embedded in $\Gr(k,n)$ \cite{Arkani-Hamed:2013kca,Arkani-Hamed:2013jha}. Moreover, integrals associated to the BCFW diagrams define a volume form on the Amplituhedron, \hlfix{which is useful for ...}{we need a vague one-line description here}. Since then, there has been significant work to understand the volume form and its physical implications \cite{UnwindingAmplituhedron,galashinlam}.

Meanwhile, a different body of work has studied SYM $N=4$ theory from the point of view of Feynman integrals in twistor space \cite{Adamo:2011pv,Adamo:2012xe,  Boels:2007qn, Bullimore:2010pj}. These integrals are calculated via holomorphic Wilson loops. As with the Amplituhedron, each Wilson loop represents the sum of a family of $N^kMHV$ Feynman diagrams, and adding up the corresponding sums of Feynman integrals associates a volume form to \hlfix{space}{I don't know if this was supposed to say ``space'' or ``this space'', and if ``this space'' I don't know which space it's referring to.} \cite{Hodges:2012ym}. The first author and Amat uncovered a connection between the Feynman integrals developed in this literature and positroids, defined  as matroids that can be realized as an element of $\Gr(k,n)$ \cite{wilsonloop}, while other work has shown a connection between the space parametrized by Wilson loop diagrams and the Amplituhedron \cite{casestudy, Amplituhedronsquared}.

In the above work and elsewhere, it is clear that while the space parametrized by the Wilson loop diagrams is related to the Amplituhedron, they are not the same space. For instance, it is conjectured that the Amplituhedron is orientable, whereas the first author and Marcott show in \cite{non-orientable} that the space parametrized by the Wilson loop diagrams can be seen as a non-orientable vector bundle over some submanifold of $\Gr(k,n)$. Furthermore, the cells associated to the Wilson loop diagrams are different from the cells associated to BCFW diagrams, as one sees from counting arguments in comparing the results of \cite{AmplituhedronDecomposition} and \cite{casestudy}. This is not too alarming, as \hlfix{one expects}{can we call this a conjecture?} the space parametrized by the Wilson loop diagrams to project down to something close to the Amplituhedron \cite{wilsonloop, Amplituhedronsquared}.

This paper continues this work by identifying which positroid cells are associated to which Wilson loop diagram: specifically, we give an algorithm to read off the Grassmann necklace of the associated positroid cell from the Wilson loop diagram. We also show that the positroid cells associated to Wilson loop diagrams are all $3k$ dimensional; by contrast, the BCFW cells are all $4k$ dimensional \hlfix{[ref]}{someone actually proved this, right?}. Again, this is not too alarming, as it was shown in \cite{non-orientable} that the actual geometric space parametrized by the Wilson loop diagrams is a non-orientable $k$-dimensional bundle over this $3k$ dimensional manifold in $\Gr(k,n)$.

As noted above, the sum of the Feynman integrals associated to Wilson loop diagrams gives the volume of \hlfix{the space}{help?}; if this volume is to be finite, then any singularities appearing in individual integrals must cancel out in the overall sum. This has been verified explicitly in a few cases \cite{casestudy, HeslopStewart}, but it is hard to prove in general due to the fact that the relationship between Wilson loop diagrams and $3k$ dimensional positroid cells of $\Gr(k,n)$ is neither one-to-one nor onto. In this paper, we give a characterization of the factors appearing in the denominator of these Feynman integrals in terms of the Grassmann necklace of the associated positroid, thus providing a much simpler and more tractable link between the singularities and the diagrams.

% when computing a finite volume to a geometric object defined by the Wilson loop diagrams, one must ensure that any singularities cancel. This has been verified explicitly in a few cases \cite{casestudy, HeslopStewart}, but it has been hard to prove in general due to the fact that the relationship between Wilson loop diagrams and $3k$ dimensional positroid cells of $\Gr(k,n)$ is neither one-to-one nor onto. In this paper, we give a characterisation of the denominator of the Feynman integral defined by a Wilson loop diagram in terms of the algorithm that associated the same diagram to a Grassman Necklace.

@@

In Section~\ref{section general background}, we summarize the required background for Wilson loop diagrams (subsection~\ref{section WLD background}), the theory of positroids in $\Gr(k,n)$ and their characterization in terms of Grassmann necklaces and Le diagrams (subsection~\ref{sec:positroid background}), and the link between Wilson loop diagrams and positroids (subsection~\ref{sec:WLD as positroids}).

In Section~\ref{sec GN algorithm} we state an algorithm for constructing the Grassmann necklace of the positroid associated to a Wilson loop diagram (Algorithm~\ref{alg:put GN on WLD}) and prove its correctness (Theorem~\ref{res:alg gives GN}). While it was already possible to construct the Grassmann necklace of a Wilson loop diagram via existing bijections between some of the various combinatorial objects that index the positroid cells, this process was convoluted and involved multiple steps; our algorithm allows us to track directly how each propagator in the Wilson loop diagram contributes to each term in the Grassmann necklace. Many of the technical lemmas required to prove Theorem~\ref{res:alg gives GN} continue to be important in subsequent sections: in particular we have Lemma~\ref{lem sian}, which identifies certain configurations of propagators which must appear in any admissible Wilson loop diagram, and Lemma~\ref{vertex cyclic int lem}, which shows that the pattern of values that a propagator contributes to each term of the Grassmann necklace is both simple and predictable.

In Section~\ref{sec dim} we examine the dimension of positroids associated to Wilson loop diagrams, and show that the dimension is always equal to three times the number of propagators in the diagram (Theorem~\ref{thm dim}).  While this has been proved by Marcott in \hlfix{[ref]}{Did this get published yet?} using geometric methods, our proof is constructive and explicitly relates the position of the plusses in the Le diagram to the propagators in the Wilson loop diagram.

Finally in Section~\ref{sec poles} we characterize the denominator of the integral of a Wilson loop diagram in terms of how propagators contribute to the Grassmann necklace.  This characterization is given explicitly by Algorithm~\ref{alg WLD to denom via GN} and Theorem~\ref{thm denom}.

@@

@@

Generally speaking the propagators in a Wilson loop diagram are undirected, so $(i,j)$ and $(j,i)$ represent the same propagator. If we need to impose a direction on a propagator $p = (i,j)$, for example in order to consider the region ``inside'' or ``outside'' of $p$, we write $(p,<_i)$ to indicate that $p$ is directed from edge $i$ to edge $j$, and $(p,<_j)$ for the opposite direction.

We also impose the convention that all vertices should always be interpreted in terms of the cyclic order on $V$; for example, if $V = [n]$ then $(n,n+3)$ means the propagator connecting edges $3$ and $n$. This viewpoint is beneficial for us since many of our proofs rely on restricting our attention to a region of the diagram bounded by a given propagator, or to short propagators such as $(i,i+2)$, and these behave the same regardless of whether we cross the ``first'' vertex or not; however, we emphasize that this is different to the standard convention in the physics literature, where propagators are written $(i,j)$ with $1 \leq i<j \leq n$.

@@

For our purposes, a {\bf positroid} is a matroid $M$ (with ground set $[n]$ and bases $\cB \subseteq \binom{[n]}{k}$) which can be represented by an element of the positive Grassmannian $\Gr(k,n)$. In other words, there exists a full-rank $k\times n$ real matrix whose $k\times k$ minors are all nonnegative and such that the minor $[[k]|J]$ is positive if and only if $J \in \cB$.

We summarize here only the subset of positroid theory that we require in this paper; the interested reader is referred to \cite{Postnikov} for more details. Postnikov shows in \cite{Postnikov} that the positroids of $\Gr(k,n)$ are indexed by many different collections of objects, each with their own advantages and disadvantages. The two most suited to our purposes are Grassmann necklaces and Le diagrams, which we introduce below. In order to do so, we first need some preliminary definitions.

@@

By \cite[Theorem 17.1]{Postnikov}, the Grassmann necklaces of type $(k,n)$ are in 1-1 correspondence with the positroid cells in $\Gr(k,n)$. Each term $I_i$ is simply the minimal (with respect to the $<_i$-lex order) basis of the positroid. A characterization of all bases of the positroid in terms of the Grassmann necklace and the Gale order was given by Oh in \cite[Theorem 8]{Oh}: if $(I_1, \dots, I_n)$ is the Grassmann necklace associated to a positroid $M = ([n],\cB)$, then the bases of $M$ are exactly
\[\cB = \left\{J \in \binom{[n]}{k}\ |\ I_i \gale{i} J \ \forall i \in [n]\right\}.\]
Thus the Grassmann necklace is well suited to testing whether a given $k$-set is a basis for $M$, and for generating a list of all bases of $M$.

@@

An algorithm for constructing the Le diagram associated to a Grassmann necklace was given by Agarwala and Fryer in \cite{reversingOh}. Since we will make use of this algorithm in Section~\ref{sec dim} below, we summarize the process here.
\begin{algorithm}\label{alg:GN to Le} \ \cite[Algorithm 2]{reversingOh}
Let $(I_1,\dots,I_n)$ be a Grassmann necklace of type $(k,n)$. Within a $k \times(n-k)$ square, draw the Young diagram whose rows are labelled by $I_1$ (as per the convention above).

@@

\begin{thm} \label{thm WLD defines matroid} \cite[Theorem 3.6]{wilsonloop} The independent sets of the matroid $M(W)$ associated to an admissible Wilson loop diagram $W = (\cP,[n])$ and realized by $C(W)$ are exactly those subsets $V \subseteq [n]$ such that $\nexists U\subseteq V$ satisfying $|\Prop(U)| < |U|$. \end{thm}
In other words, the independent sets of $M(W)$ correspond to sets of vertices in $W$ in which no subset supports fewer propagators than the vertices it contains.

By studying $M(W)$ in terms of the behavior of the propagators in $W$, Agarwala and Amat also showed that $M(W)$ is in fact a positroid whenever $W$ is admissible \cite[Corollary 3.39]{wilsonloop}. In theory, this allows us to leverage geometric and combinatorial results about $\Gr(k,n)$ to study Wilson loop diagrams.

However, in order to identify the positroid associated to a given Wilson loop diagram $W$ we first have to construct $C(W)$, then make a list of all its non-zero $k\times k$ minors and extract the Le diagram or Grassmann necklace via the methods described in Section~\ref{sec:positroid background}. Alternatively, one could obtain the decorated permutation of the positroid from $C(W)$ as in \hlfix{[ref]}{Sian needs to go find this reference} and then convert it into a Grassmann necklace as described in \cite[Section 16]{Postnikov}.

In \cite{casestudy}, Agarwala and Fryer apply this process in the smallest non-trivial case: admissible Wilson loop diagrams with two propagators on six vertices. Even in this simple case, we see that the mapping from admissible WLD to positroid cells is neither one-to-one nor onto, and the relationship between the original Wilson loop diagram and the resulting positroid is obscured to the point of invisibility.

Our first goal is therefore to find a better method of obtaining the positroid associated to a given Wilson loop diagram; this is the focus of the next section.

@@

\begin{dfn}\label{props inside p}
Let $W = (\cP, [n])$ be a weakly admissible Wilson loop diagram, with $p \in \cP$ supported on edges $i$ and $j$. Write $(p, <_i)$ to represent the same propagator directed from $i$ to $j$. Then the sets of propagators {\bf inside} and {\bf outside} of $(p,<_i)$ are defined to be
\begin{gather*}\cP_{in}(p,<_i) = \{ (k,l) \in \cP \ |\ i \leq_i k <_i l \leq_i j \}, \\
\cP_{out}(p,<_i) = \cP\setminus \cP_{in}(p,<_i),
\end{gather*}
respectively.
\end{dfn}
